---
title: 'PK/PD Modeling'
author: "Anton Sun"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 5
    number_sections: true
    latex_engine: tinytex
  word_document: default
  pdf_document:
    toc: true
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library( CyneRgy )
```

# Introduction
The following example demonstrates the integration capabilities of PK/PD modeling into East Horizon using R functions. In the example, we are performing a two arm trial, including a control arm and an experimental treatment arm. In our function, we are utilizing a one-compartment model with first-order absorption and first-order elimination to output plasma concentration.

Once CyneRgy is installed, you can load this example in RStudio with the following commands:
```{r, eval=FALSE}
CyneRgy::RunExample( "PKPDExample" )
```
**RStudio Project File**: [PKPDExample.Rproj](https://github.com/Cytel-Inc/CyneRgy/blob/f31e872bad8b241627b251789a0b6e0e40d05ac9/inst/Examples/PKPDExample/PKPDExample.Rproj)

In the [R directory of this example](https://github.com/Cytel-Inc/CyneRgy/tree/f31e872bad8b241627b251789a0b6e0e40d05ac9/inst/Examples/PKPDExample/RCode) you will find the following R files:

1. [GenerateDrugConcentration.R](https://github.com/Cytel-Inc/CyneRgy/blob/f31e872bad8b241627b251789a0b6e0e40d05ac9/inst/Examples/PKPDExample/RCode/GenerateDrugConcentration.R)

2. [ImprovedGenerateResponseEmaxModel.R](https://github.com/Cytel-Inc/CyneRgy/blob/f31e872bad8b241627b251789a0b6e0e40d05ac9/inst/Examples/PKPDExample/RCode/ImprovedGenerateResponseEmaxModel.R)

3. [PKPDResponseGeneration.R](https://github.com/Cytel-Inc/CyneRgy/blob/f31e872bad8b241627b251789a0b6e0e40d05ac9/inst/Examples/PKPDExample/RCode/PKPDResponseGeneration.R)

<!-- PK:
     Show equations and variables
     Initial conditions of A1 and A2
     Use cases
     
     PD:
     EMaxMOdel
     - Direct Effect
     - Indirect Effect
     
     ImprovedGenerateResponseEmax