---
title: "2-Arm, Single Endpoint - Simulate Patient Dropout"
author: "Shubham Lahoti"
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0">These examples are related to the <strong>Integration Point: Dropout</strong>.
  <a href="IntegrationPointDropout.html" class="alert-link">Click here for more information about this integration point.</a></p>
</div>

## Introduction

The following examples illustrate how to customize the dropout distribution in East Horizon or East using R functions. Patients may drop out of a trial for various reasons, including safety concerns, treatment burden, or non-trial-related factors. In some cases, dropout rates can reach 30%, particularly if the drug has adverse side effects. Incorporating dropout probabilities or hazard rates is crucial during data generation and can significantly impact subsequent analysis. In all examples, we assume a trial design consisting of a control group and an experimental treatment.

Once CyneRgy is installed, you can load this example in RStudio with the following commands:
```{r, eval=FALSE}
CyneRgy::RunExample( "2ArmPatientDropout" )
```

Running the command above will load the RStudio project in RStudio. 

**RStudio Project File**: [2ArmPatientDropout.Rproj](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/2ArmPatientDropout.Rproj)

In the [R directory of this example](https://github.com/Cytel-Inc/CyneRgy/tree/main/inst/Examples/2ArmPatientDropout/R) you will find the following R files:

1. [GenerateCensoringUsingBinomialProportion.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateCensoringUsingBinomialProportion.R) - Contains a function named *GenerateCensoringUsingBinomialProportion* to demonstrate the R code necessary for Example 1 as described below.

2. [GenerateDropoutTimeForSurvival.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateDropoutTimeForSurvival.R) - Contains a function named *GenerateDropoutTimeForSurvival* to demonstrate the R code necessary for Example 2 as described below.

3. [GenerateDropoutTimeForRM](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateDropoutTimeForRM) - Contains a function named *GenerateDropoutTimeForRM* to demonstrate the R code necessary for Example 3 as described below.

# Example 1 - Dropout Using Binomial Proportion

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateCensoringUsingBinomialProportion.R" class="alert-link">GenerateCensoringUsingBinomialProportion.R.R</a></p>
</div>

The R function *GenerateCensoringUsingBinomialProportion* generates the censoring indicator using the dropout probability specified in East Horizon (*ProbDrop* in the script, *Probability of Dropout* in East Horizon) for continuous or binary outcome. This function applies the binomial distribution using the `rbinom()` function in R to determine dropout status. In this case, the dropout probability is the same across both treatment groups, but this can be customized using user-defined parameters. 

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Probability of Dropout|Response Card|$ProbDrop$|

Steps:  

1. Let $pd$ be the dropout probability.  
2. Draw a random sample from a Bernoulli distribution with $p = 1 - pd$, i.e., $Binomial(1, pd)$, of size $n = NumSub$.  
3. The generated sample serves as a censoring indicator:  
   - **1** indicates a patient who completes the trial (does not drop out).  
   - **0** indicates a patient who drops out (non-completer).  
   
The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Dropout" = c(
            "Load dropout probability", 
            "Loop through patients", 
            "Simulate whether patient drops out using binomial distribution", 
            "Return censoring indicators"
        )
    )
)
```

# Example 2 - Dropout Time For Survival

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateDropoutTimeForSurvival.R" class="alert-link">GenerateDropoutTimeForSurvival.R</a></p>
</div>

The function *GenerateDropoutTimeForSurvival* generates dropout times for a two-arm survival design. In this case, dropout information can be specified separately for each arm (in East Horizon, *Control* and *Treatment* in the *Dropout Rate* tab), and dropout times are drawn from an exponential distribution accordingly. The number of periods is fixed at 1.

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Probability of Dropout: Control|Response Card, Dropout Rate tab|$DropParam[ 1 ]$|
|Probability of Dropout: Treatment|Response Card, Dropout Rate tab|$DropParam[ 2 ]$|

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Dropout" = c(
            "Load dropout probabilities", 
            "Calculate Hazard rates from probabilities",
            "Loop through patients", 
            "Simulate whether patient drops out using exponential distribution", 
            "Return dropout times"
        )
    )
)
```

# Example 3 - Dropout Time for Repeated Measures

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmPatientDropout/R/GenerateDropoutTimeForRM.R" class="alert-link">GenerateDropoutTimeForRM.R</a></p>
</div>

The function *GenerateDropoutTimeForRM* generates dropout times for a continuous outcome with repeated measures design. In this case, dropout information can be specified separately for each arm (in East Horizon, *Control* and *Treatment* in the *Dropout Rate* tab), and dropout times are drawn from an exponential distribution accordingly. The *Input Method* in East Horizon has to be set to *Cumulative Probability of Dropout by Time*.

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Probability of Dropout: Control|Response Card, Dropout Rate tab|$DropParamControl$|
|Probability of Dropout: Treatment|Response Card, Dropout Rate tab|$DropParamtrt$|

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Dropout" = c(
            "Load dropout probabilities", 
            "Calculate dropout rates from probabilities",
            "Loop through patients", 
            "Simulate whether patient drops out using exponential distribution", 
            "Return dropout times"
        )
    )
)
```
