---
title: "2-Arm, Binary Outcome - Analysis"
author: "J. Kyle Wathen and Gabriel Potvin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: default
  word_document: default
---

<div class="alert alert-primary" role="alert">
<p style="margin-bottom:0">This example is related to the <a href="IntegrationPointAnalysisBinary.html" class="alert-link">**Integration Point: Analysis - Binary Outcome**</a>. Click the link for setup instructions, variable details, and additional information about this integration point.</p></div>

<div class="alert alert-danger" role="alert">
To try this example, create a new project in East Horizon using the following configuration:
<ul style="margin-bottom:0">
<li>**Study objective:** Two Arm Confirmatory</li>
<li>**Number of endpoints:** Single Endpoint</li>
<li>**Endpoint type:** Binary Outcome</li>
<li>**Task:** Any</li>
<li>**Statistical Design:**</li>
<ul style="margin-bottom:0">
<li>*Fixed Sample* or *Group Sequential* for Examples 1, 2, 3, 4</li>
<li>*Group Sequential with Sample Size Re-Estimation* for Example 5</li>
</ul>
</ul></div>

<div class="alert alert-warning" role="alert">
<p style="margin-bottom:0">**Note:** Examples 1, 2, 3, and 4 are compatible with both *Fixed Sample* and *Group Sequential* statistical designs. The R code automatically detects whether interim look information (*LookInfo*) is available and adjusts the analysis parameters accordingly.</p>
</div>

# Introduction

The following examples illustrate how to integrate new analysis capabilities into East Horizon or East using R functions in the context of a two-arm trial. In each example, the trial design includes a standard-of-care control arm and an experimental treatment arm, with patient outcomes modeled as binary data. The design includes two interim analyses (IA) and one final analysis (FA). At each IA, an analysis is conducted which may lead to early stopping for efficacy or futility, depending on the predefined design criteria.

Once CyneRgy is installed, you can load this example in RStudio with the following commands:
```{r, eval=FALSE}
CyneRgy::RunExample( "2ArmBinaryOutcomeAnalysis" )
```

Running the command above will load the RStudio project in RStudio. 

**East Workbook**: [2ArmBinaryOutcomeAnalysis.cywx](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/2ArmBinaryOutcomeAnalysis.cywx)

**RStudio Project File**: [2ArmBinaryOutcomeAnalysis.Rproj](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/2ArmBinaryOutcomeAnalysis.Rproj)

In the [R directory of this example](https://github.com/Cytel-Inc/CyneRgy/tree/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R) you will find the following R files:

1.  [AnalyzeUsingEastManualFormula.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingEastManualFormula.R) - Contains a function named *AnalyzeUsingEastManualFormula* that computes the two-sample Z-statistic for a binary endpoint using Formula 24.2 from the East manual.

2.  [AnalyzeUsingPropTest.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingPropTest.R) - Contains a function named *AnalyzeUsingPropTest* that performs the analysis using R's built-in `prop.test()` function.

3.  [AnalyzeUsingPropLimitsOfCI.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingPropLimitsOfCI.R) - Contains a function named *AnalyzeUsingPropLimitsOfCI* that bases interim and final decisions on confidence interval limits rather than East Horizon-provided boundaries. The function uses user-defined parameters (MAV, TV, and confidence level) to implement a Go / No-Go decision framework.

4.  [AnalyzeUsingBetaBinomial.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingBetaBinomial.R) - Contains a function named *AnalyzeUsingBetaBinomial* that performs a Bayesian analysis for binary outcomes using Beta-Binomial conjugate priors.

5.  [AnalyzeBinarySSR.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeBinarySSR.R) - Contains a function named *AnalyzeBinarySSR* that extends the standard binary analysis to support conditional power–based sample size re-estimation (SSR) in a group sequential design.

In addition, if you would like to experiment with these examples to and would like some code to help you get started we have provided fill-in-the-blank type code files in the [FillInTheBlankR directory](https://github.com/Cytel-Inc/CyneRgy/tree/main/inst/Examples/2ArmBinaryOutcomeAnalysis/FillInTheBlankR).  

# Example 1 - Using Formula 24.2 from the East manual

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingEastManualFormula.R" class="alert-link">AnalyzeUsingEastManualFormula.R</a></p>
</div>

In this example, the analysis is customized by replacing the default method with a user-defined calculation based on formulas of the **Chapter 24: Binomial Superiority Two‐Sample** from the East manual (24.2).

**Test Statistic:**

$$Z_j = \frac{\hat{\pi}_{tj} - \hat{\pi}_{cj}}{\sqrt{\hat{\pi}_j \cdot (1 - \hat{\pi}_j) \cdot ( \frac{1}{n_{tj}} + \frac{1}{n_{cj}} )}}$$

Where:

- $\hat{\pi}_{tj}$ and $\hat{\pi}_{cj}$ are the observed response rate at the jth interim look in the treatment and control groups, respectively.
- $n_{tj}$ and $n_{cj}$ are the numbers of patients at the jth interim look in the treatment and control groups, respectively.
- $\hat{\pi}_{j} = \frac{n_{tj} \cdot \hat{\pi}_{tj} + n_{cj} \cdot \hat{\pi}_{cj}}{n_{tj} + n_{cj}}$ is the pooled response rate estimate.

The objective is to demonstrate a straightforward way to modify both the analysis and decision-making process. The computed test statistic is compared to the efficacy boundary provided by East Horizon or East as input. This example does not include a futility rule and does not use any user-defined parameters.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load responses from East Horizon", 
            "Calculate response rates and estimates", 
            "Calculate Z-statistic (formula)", 
            "Return decision using EH boundary"
        )
    )
)
```

# Example 2 - Using the `prop.test()` Function in R

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingPropTest.R" class="alert-link">AnalyzeUsingPropTest.R</a></p>
</div>

This example utilizes the base R `prop.test()` function to perform the interim and final analyses. The p-value is used to compute the Z-statistic, which is compared against the efficacy boundary provided by East Horizon or East. Like Example 1, this example does not include a futility rule and does not use any user-defined parameters.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load responses from East Horizon", 
            "Run proportions test (treatment > control)", 
            "Extract Z-statistic from p-value", 
            "Return decision using EH boundary"
        )
    )
)
```

# Example 3 - Utilization of Confidence Interval Limits for Go/No-Go Decision-Making

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingPropLimitsOfCI.R" class="alert-link">AnalyzeUsingPropLimitsOfCI.R</a></p>
</div>

In many Phase II trials, Go/No-Go decisions are made based on whether a treatment shows sufficient promise to justify further development. These decisions are often guided by two key thresholds:

- **Minimum Acceptable Value (MAV)**: The smallest treatment effect considered meaningful.
- **Target Value (TV)**: A highly desirable treatment effect based on clinical or strategic considerations.

This example demonstrates how to approximate probabilistic decision-making using frequentist confidence intervals (CIs), ignoring the boundaries provided by East Horizon or East in favor of a CI-based logic. We use the function `prop.test()` from base R to analyze the data and compute the desired confidence intervals. If the treatment difference is likely to exceed the MAV, a Go decision is made. If not, and it is unlikely to exceed the TV, a No-Go decision is made. Specifically:

**At Interim Analysis**

- Let LL and UL be the lower and upper limits of the confidence interval for the treatment effect.
- If $LL > \text{MAV} \rightarrow \text{Go}$
- If $UL < \text{TV} \rightarrow \text{No-Go}$
- Otherwise $\rightarrow$ Continue to the next analysis

**At Final Analysis**

- If $LL > \text{MAV} \rightarrow \text{Go}$
- Otherwise $\rightarrow$ No-Go

Refer to the table below for the definitions of the user-defined parameters used in this example.

|**User parameter**|**Definition**|
|--|------|
|**dMAV**|Minimum Acceptable Value: the smallest treatment effect considered clinically meaningful to warrant further development.|
|**dTV**|Target Value: the desired treatment effect that would represent a strong clinical benefit or strategic advantage.|
|**dConfLevel**|Confidence Level: the level of confidence used to construct the confidence interval for Go/No-Go decision-making (e.g., 0.80 for an 80% CI).|

**Note:** In this example, the boundary information that is computed in East Horizon or East is ignored. User-defined parameters and the function `prop.test()` from base R are used to analyze the data and compute the desired confidence intervals.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load responses from East Horizon",
            "Load MAV, TV, and confidence level",
            "Run proportions test (treatment > control)", 
            "Calculate CI lower and upper limits", 
            "Return decision using CI thresholds"
        )
    )
)
```

# Example 4 - Bayesian Analysis

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeUsingBetaBinomial.R" class="alert-link">AnalyzeUsingBetaBinomial.R</a></p>
</div>

This example illustrates how prior data on patients can be incorporated directly into a Bayesian model. The user can specify Beta prior distributions for both the standard of care and experimental arms, as well as set decision thresholds for efficacy and futility. Refer to the table below for the definitions and values of the user-defined parameters used in this example.

|**User parameter**|**Definition**|**Value**|
|--|------|--|
|**dAlphaCtrl**|Alpha parameter of the Beta prior for the control group.|10|
|**dBetaCtrl**|Beta parameter of the Beta prior for the control group.|40|
|**dAlphaExp**|Alpha parameter of the Beta prior for the treatment group.|0.2|
|**dBetaExp**|Beta parameter of the Beta prior for the treatment group.|0.8|
|**dUpperCutoffEfficacy**|Upper threshold for declaring efficacy (between 0 and 1). If the posterior probability exceeds this value, efficacy is declared.|0.95|
|**dLowerCutoffForFutility**|Lower threshold for declaring futility (between 0 and 1). If the posterior probability falls below this value, futility is declared.|0.10|

In this example and with the values specified above, the prior data includes 10 responders and 40 non-responders. Let $\pi_S$ denote the response rate under the standard of care, and $\pi_E$ the response rate under the experimental treatment. The prior distributions are defined as:

- $\pi_S \sim \text{Beta}(10, 40)$: implies a prior mean of 20%, equivalent to observing 50 prior patients with 10 responses.
- $\pi_E \sim \text{Beta}(0.2, 0.8)$: implies a prior mean of 20%, but corresponds to just 1 prior patient, reflecting a weakly informative prior.

The Beta distribution is conjugate for binary (binomial) data, making it convenient for Bayesian updating. Suppose $N_{\text{Ctrl}}$ and $N_{\text{Exp}}$ patients have enrolled in the control and experimental arms, respectively, and we observe at an interim analysis:

| Treatment | Number of Responses | Number of Treatment Failures |
|-----------|---------------------|------------------------------|
|    Control   | 	$X_{\text{Ctrl}}$          | $N_{\text{Ctrl}} - X_{\text{Ctrl}}$        |
|    Experimental    | $X_{\text{Exp}}$           | $N_{\text{Exp}} - X_{\text{Exp}}$          |


The posterior distributions for the response rates become:

$$\pi_{\text{Ctrl}} \mid X_{\text{Ctrl}}, N_{\text{Ctrl}} \sim \text{Beta}(10 + X_{\text{Ctrl}},\ 40 + N_{\text{Ctrl}} - X_{\text{Ctrl}})$$

$$\pi_{\text{Exp}} \mid X_{\text{Exp}}, N_{\text{Exp}} \sim \text{Beta}(0.2 + X_{\text{Exp}},\ 0.8 + N_{\text{Exp}} - X_{\text{Exp}})$$

For analysis and decision-making, the posterior distributions are used to determine whether to stop the trial early. The trial is stopped for efficacy if there is strong evidence that the response rate in the experimental group is higher than in the control group, and for futility if there is little evidence to support a benefit of the experimental treatment over the standard of care. The primary quantity of interest is the posterior probability that the experimental treatment has a higher response rate than the standard of care:

$$\rho = \Pr(\pi_{\text{Exp}} > \pi_{\text{Ctrl}} \mid X_{\text{Ctrl}}, N_{\text{Ctrl}}, X_{\text{Exp}}, N_{\text{Exp}})$$
Specifically,

**At Interim Analysis**

- If $\rho > 0.975 \rightarrow$ Make an early efficacy (Go) decision.
- If $\rho < 0.10 \rightarrow$ Make an early futility (No-Go) decision.

**At Final Analysis**

- If $\rho > 0.95 \rightarrow$ Make an efficacy decision.
- Otherwise $\rightarrow$ Make a futility decision.

Note that the above values can be user-specified. This example demonstrates how incorporating prior information on $\pi_S$ can influence decision-making during a trial. However, it is essential to conduct sensitivity analyses to assess how the choice of priors affects operating characteristics and overall conclusions.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load responses from East Horizon",
            "Load prior parameters and thresholds",
            "Compute posterior parameters using Beta-Binomial model", 
            "Compute posterior probability that experimental > control", 
            "Return decision using thresholds"
        )
    )
)
```

# Example 5 - Analysis With Sample Size Re-Estimation

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmBinaryOutcomeAnalysis/R/AnalyzeBinarySSR.R" class="alert-link">AnalyzeBinarySSR.R</a></p>
</div>

This example illustrates how to customize the analysis when using `Statistical Design = Group Sequential with Sample Size Re-Estimation` in East Horizon. The key difference from the previous examples is that the analysis function is responsible not only for producing a test statistic and interim decision, but also for computing a re-estimated total number of completers based on interim results. This is achieved using a conditional power–based SSR approach, which allows the trial to increase its planned number of completers when interim results fall within a "promising zone".

**Test Statistic:**

A pooled-variance Z-statistic is computed based on the observed difference in response rates between the experimental and control arms:

$$Z = \frac{p_{exp} - p_{ctrl}}{\sqrt{\hat{p} \cdot (1 - \hat{p}) \cdot ( \frac{1}{n_{exp}} + \frac{1}{n_{ctrl}} )}}$$

where $\hat{p}$ is the pooled response rate across both arms.

**Conditional Power:**

Using the observed Z-statistic, the information fraction at the current look, and the efficacy boundary provided by East Horizon, the conditional power is computed. This represents the probability of eventually crossing the final efficacy boundary, given the interim data.

**Sample Size Re-Estimation (SSR):**

Based on the computed conditional power, the total number of completers is re-estimated using one of two SSR rules:

|Condition|Rule|
|----|----|
|The SSR function scale for promising zone selected in East Horizon is "Continuous" (`AdaptInfo$SSRFuncScale == 0`)|If the conditional power falls between the lower and upper promising-zone thresholds selected in East Horizon (`AdaptInfo$PromZoneMin` and `AdaptInfo$PromZoneMax`), the maximum number of completers is multiplied by the maximum sample size multiplier input specified in East Horizon (`AdaptInfo$MaxSSMultInp$MaxSSMult`).|
|The SSR function scale for promising zone selected in East Horizon is "Step" (`AdaptInfo$SSRFuncScale == 1`)|The conditional power is mapped to intervals specified in East Horizon (`AdaptInfo$MaxSSMultInp$From`, `AdaptInfo$MaxSSMultInp$To`), each associated with a specific sample size multiplier specified in East Horizon (`AdaptInfo$MaxSSMultInp$MaxSSMult`).|

If the conditional power falls outside the promising zone, the original planned number of completers is retained.

Finally, the Z-statistic is compared against the efficacy boundary provided by East Horizon. At interim looks, crossing the boundary results in an early efficacy decision, while at the final look a lack of boundary crossing results in a futility decision. Both the decision and the re-estimated total number of completers (`ReEstCompleters`) are returned to East Horizon.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load responses and arrival times from EH",
            "Compute Z-statistic and observed treatment effect",
            "Compute conditional power using efficacy boundary",
            "Re-estimate total completers",
            "Return decision and re-estimated number of completers"
        )
    )
)
```
