---
title: 'MMRM Analysis'
author: "Jacob Wathen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 5
    number_sections: true
    latex_engine: tinytex
  word_document: default
  pdf_document:
    toc: true
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library( CyneRgy )
```

# Abstract

This document summarizes the goals, background, methods, simulations, results, and takeaways from the CyneRgy MMRM R development work and companion East Horizon simulations completed. It distills the content of the project into a concise narrative.

> **Scope note.** This write‑up focuses on Mixed Models for Repeated Measures (MMRM) analyses for continuous outcomes over time in schizophrenia trials using PANSS as a primary endpoint, plus comparable simulations performed in East Horizon.


# Introduction

- This example will cover the setup required to be customized to match an expected clinical trial of lumateperone tosylate on  Schizophrenia.

Our primary outcome in this clinical trial is a change from baseline to follow-up in the Positive and Negative Syndrome Scale (PANNS), it assess schizophrenia on 3 criteria across 30 items. These criteria are positive symptoms (7 items), negative symptoms (7 items), and general psychopathology (16 items). Each of these is rated on a 7-point scale (1 = absent, 7 = extreme). The lower values representing less symptoms of schizophrenia the range being from 30 to 210. The difference from baseline to follow-up must be a change in 6 points to show efficacy after 28 days of treatment. During this 28-day period each individual was measured again after each week.

# Required Files

Before running anything in East Horizon, load the Schizophrenia example by installing CyneRgy and running this command into RStudio:

```{r, eval=FALSE}
CyneRgy::RunExample( "Schizophreniatrial" )
```

# East Horizon Simulation Setup

To recreate this example for yourself, create a new project in East Horizon using the Continuous endpoint type. Next create an input set using the explore option.


- **Designs**: MMRM fixed‑sample (baseline N ~266), group‑sequential MMRM, and null fixed‑sample difference‑in‑means (all at Type I error 0.025).
- **Responses**: PANSS change with SD = 15 (baseline), plus SD = 10 and SD = 20 variants; a **null** setting used control‑like outcomes for the treatment arm.
- **Dropout & Enrollment**: Shared dropout card (from publication assumptions); total N grid from 200 to 360 in steps of 50 for MMRM.

# Results (Selected Highlights)

- **Target power**: ≈90% power achieved with ~190 completers at SD = 15 (implies larger randomized N depending on dropout).
- **Sensitivity**: Power changed as SD and mean difference were varied (SD ↑ → power ↓; SD ↓ → power ↑).
- **Null behavior (R MMRM)**: Type I error around 2.1–3.6%, close to nominal 2.5%.
- **Group‑sequential caution (EH)**: Using native East Horizon group‑sequential analysis inflated the null rejection rate (e.g., ~17.6% vs. 2.5% expected). Prefer validated MMRM implementations or investigate analysis settings.

# Discussion

- **Model choice**: MMRM is well‑suited for longitudinal continuous outcomes with missing‑at‑random assumptions; it uses all available post‑baseline data without LOCF and models the mean profile over time.
- **Practicalities**: Reshaping (wide→long), baseline adjustment, visit factorization, and correlation structure (e.g., AR(1)) are standard components.
- **Visualization**: Trial‑level mean trajectories with 95% CIs and optional patient‑level plots help QC data generation.
- **Platform differences**: Ensure parity in estimands, covariance structures, and decision rules between R and East Horizon; verify Type I error control under the null when using group‑sequential options.


