---
title: '2-Arm, Time-To-Event Outcome - Analysis'
author: "Anoop Singh Rawat and Shubham Lahoti"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: default
  word_document: default
---

<div class="alert alert-primary" role="alert">
<p style="margin-bottom:0">This example is related to the <a href="IntegrationPointAnalysisTimeToEvent.html" class="alert-link">**Integration Point: Analysis - Time-to-Event Outcome**</a>. Click the link for setup instructions, variable details, and additional information about this integration point.</p></div>

<div class="alert alert-danger" role="alert">
To try this example, create a new project in East Horizon using the following configuration:
<ul style="margin-bottom:0">
<li>**Study objective:** Two Arm Confirmatory</li>
<li>**Number of endpoints:** Single Endpoint</li>
<li>**Endpoint type:** Time-to-Event Outcome</li>
<li>**Task:** Any</li>
</ul></div>

<div class="alert alert-warning" role="alert">
<p style="margin-bottom:0">**Note:** This example is compatible with both Fixed Sample and Group Sequential statistical designs. The R code automatically detects whether interim look information (*LookInfo*) is available and adjusts the analysis parameters accordingly.</p>
</div>

# Introduction

The following examples illustrate how to integrate new analysis capabilities into East Horizon or East using R functions in the context of a two-arm trial. In each example, the trial design includes a standard-of-care control arm and an experimental treatment arm, with patient outcomes modeled as time-to-event data. The design includes two interim analyses (IA) and one final analysis (FA). At each IA, an analysis is conducted which may lead to early stopping for efficacy or futility, depending on the predefined design criteria.

Once CyneRgy is installed, you can load this example in RStudio with the following commands:
```{r, eval=FALSE}
CyneRgy::RunExample( "2ArmTimeToEventOutcomeAnalysis" )
```

Running the command above will load the RStudio project in RStudio. 

**RStudio Project File**: [2ArmTimeToEventOutcomeAnalysis.Rproj](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/2ArmTimeToEventOutcomeAnalysis.Rproj)

In the [R directory of this example](https://github.com/Cytel-Inc/CyneRgy/tree/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R) you will find the following R files:

1. [AnalyzeUsingEastLogrankFormula.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingEastLogrankFormula.R) - Contains a function named *AnalyzeUsingEastLogrankFormula* to demonstrate the R code necessary for Example 1 as described below.
2. [AnalyzeUsingSurvivalPackage.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingSurvivalPackage.R) - Contains a function named *AnalyzeUsingSurvivalPackage* to demonstrate the R code necessary for Example 2 as described below.
3. [AnalyzeUsingHazardRatioLimitsOfCI.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingHazardRatioLimitsOfCI.R) - Contains a function named *AnalyzeUsingHazardRatioLimitsOfCI* to demonstrate the R code necessary for Example 3 as described below.


# Example 1 - Using Formulas Q.242 and Q.243 from the East manual

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingEastLogrankFormula.R" class="alert-link">AnalyzeUsingEastLogrankFormula.R</a></p>
</div>

In this example, the analysis is customized by replacing the default method with a user-defined calculation based on formulas of the **Appendix Q.8 - Survival : Two Samples** from the East manual (Q.242 and Q.243).

**Numerator of Test Statistic:**

$$\text{Num} = \sum_{i=1}^M W_i \cdot \text{Num}_i$$

With:

$$\text{Num}_i = d_{i,t} - n_{i,t} \cdot \frac{d_i}{n_i}$$

Where:

- $W_i = 1$ is the weight for the Logrank Test.
- $d_{i,t}$ and $d_{i,c}$ are the numbers of events at time $t_i$ in the treatment and control groups, respectively.
- $n_{i,t}$ and $n_{i,c}$ are the numbers of subjects at risk just before time $t_i$ in the treatment and control groups, respectively.
- $d_i = d_{i,t} + d_{i,c}$ is the total number of events at time $t_i$.
- $n_i = n_{i,c} + n_{i,c}$ is the total number of subjects at risk just before time $t_i$.

**Denominator of Test Statistic:**

$$\text{Den} = \sum_{i=1}^MW_i^2 \cdot \text{Den}_i$$

With:

$$\text{Den}_i = \frac{n_{i,t} \cdot n_{i,c} \cdot (n_i - d_i) \cdot d_i}{n_i^2 \cdot (n_i - 1)}$$

The objective is to demonstrate a straightforward way to modify both the analysis and decision-making process. The computed test statistic is compared to the efficacy boundary provided by East Horizon or East as input. This example does not include a futility rule and does not use any user-defined parameters.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load survival times from East Horizon", 
            "Fit Cox model to estimate hazard ratio", 
            "Calculate log-rank test statistic (formulas)", 
            "Return decision using EH boundary"
        )
    )
)
```

# Example 2 - Using the `survival::survdiff()` Function

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingSurvivalPackage.R" class="alert-link">AnalyzeUsingSurvivalPackage.R</a></p>
</div>

This example utilizes the `survival::survdiff()` function from the `survival` package to perform the log-rank test. The resulting chi-square statistic is used to compute the log-rank value, which is then compared to the lower efficacy boundary provided by East Horizon or East. Like Example 1, this example does not include a futility rule and does not use any user-defined parameters.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.


```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load survival times from East Horizon", 
            "Fit Cox model to estimate hazard ratio", 
            "Compute log-rank test statistic (survdiff)", 
            "Return decision using EH boundary"
        )
    )
)
```

# Example 3 - Utilization of Confidence Interval Limits for Go/No-Go Decision-Making

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/2ArmTimeToEventOutcomeAnalysis/R/AnalyzeUsingHazardRatioLimitsOfCI.R)" class="alert-link">AnalyzeUsingHazardRatioLimitsOfCI.R</a></p>
</div>

In many Phase II trials, Go/No-Go decisions are made based on whether a treatment shows sufficient promise to justify further development. These decisions are often guided by two key thresholds:

- **Minimum Acceptable Value (MAV)**: The smallest treatment effect (e.g., hazard ratio) considered meaningful.
- **Target Value (TV)**: A highly desirable treatment effect based on clinical or strategic considerations.

This example demonstrates how to approximate probabilistic decision-making using frequentist confidence intervals (CIs), ignoring the boundaries provided by East Horizon or East in favor of a CI-based logic. If the hazard ratio is likely to be below the MAV, a Go decision is made. If not, and it is unlikely to be below the TV, a No-Go decision is made.

We use the Cox proportional hazards model via the `survival::coxph()` function to estimate the log hazard ratio (log HR) and its standard error. Based on these estimates, frequentist CIs are constructed to inform decision-making. Specifically:

**At Interim Analysis**

- Let LL and UL be the lower and upper limits of the confidence interval for the log HR.
- If $UL < \log(\text{MAV}) \rightarrow \text{Go}$
- If $LL > \log(\text{TV}) \rightarrow \text{No-Go}$
- Otherwise $\rightarrow$ Continue to the next analysis

**At Final Analysis**

- If $UL < \log(\text{MAV}) \rightarrow \text{Go}$
- Otherwise $\rightarrow$ No-Go

Refer to the table below for the definitions of the user-defined parameters used in this example.

|**User parameter**|**Definition**|
|--|------|
|**dMAV**|Minimum Acceptable Value: the smallest treatment effect considered clinically meaningful to warrant further development.|
|**dTV**|Target Value: the desired treatment effect that would represent a strong clinical benefit or strategic advantage.|
|**dConfLevel**|Confidence Level: the level of confidence used to construct the confidence interval for Go/No-Go decision-making (e.g., 0.80 for an 80% CI).|

**Note:** In this example, the boundary information that is computed in East Horizon or East is ignored. User-defined parameters and the Cox model are used to analyze the data and compute the desired confidence intervals.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Analysis" = c(
            "Load survival times from East Horizon",
            "Load MAV, TV, and confidence level",
            "Fit Cox model to estimate log hazard ratio", 
            "Calculate CI lower and upper limits for log HR", 
            "Return decision using CI thresholds"
        )
    )
)
```
