---
title: 'Randomization of Subjects'
author: "Shubham Lahoti, Gabriel Potvin, Anoop Singh Rawat"
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

<div class="alert alert-primary" role="alert">
<p style="margin-bottom:0">This example is related to the <a href="IntegrationPointRandomization.html" class="alert-link">**Integration Point: Randomization**</a>. Click the link for setup instructions, variable details, and additional information about this integration point.</p></div>

<div class="alert alert-danger" role="alert">
To try this example, create a new project in East Horizon using the following configuration:
<ul style="margin-bottom:0">
<li>**Study objective:**</li>
<ul style="margin-bottom:0">
<li>Two Arm Confirmatory for Examples 1, 2, 3</li>
<li>Multiple Arm Confirmatory for Example 4</li></ul>
<li>**Number of endpoints:** Any</li>
<li>**Endpoint type:** Any</li>
<li>**Task:** Any</li>
</ul></div>

# Introduction

The following examples illustrate how to integrate new custom randomization methods and control how subjects are assigned to designated arms into East Horizon or East using R functions in the context of a two-arm trial.

Once CyneRgy is installed, you can load this example in RStudio with the following commands

```{r, eval=FALSE}
CyneRgy::RunExample( "RandomizeSubjects" )
```

Running the command above will load the RStudio project in RStudio. 

**RStudio Project File**: [RandomizeSubjects.Rproj](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/RandomizeSubjects.Rproj)

In the [R directory of this example](https://github.com/Cytel-Inc/CyneRgy/tree/main/inst/Examples/RandomizeSubjects/R) you will find the following R files:

1. [RandomizationSubjectsUsingUniformDistribution.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizationSubjectsUsingUniformDistribution.R) - Contains a function named *RandomizationSubjectsUsingUniformDistribution* to demonstrate the R code necessary for Example 1 as described below.

2. [RandomizationSubjectsUsingSampleFunctionInR.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizationSubjectsUsingSampleFunctionInR.R) - Contains a function named *RandomizationSubjectsUsingSampleFunctionInR* to demonstrate the R code necessary for Example 2 as described below.

3. [BlockRandomizationSubjectsUsingRPackage.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/BlockRandomizationSubjectsUsingRPackage.R) - Contains a function named *BlockRandomizationSubjectsUsingRPackage* to demonstrate the R code necessary for Example 3 as described below.

4. [LoadrandomizeR.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/LoadRandomizeR.R) - This file is used to install the *randomizeR* package for execution of Block Randomization in R.

5. [RandomizeSubjectsAcrossMultipleArms.R](https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizeSubjectsAcrossMultipleArms.R) - Contains a function named *RandomizeSubjectsAcrossMultipleArms* to demonstrate the R code necessary for Example 4 as described below.

# Example 1 - Randomize Subjects Usings Uniform Distribution

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizationSubjectsUsingUniformDistribution.R" class="alert-link">RandomizationSubjectsUsingUniformDistribution.R</a></p>
</div>

The R function *RandomizationSubjectsUsingUniformDistribution* randomly assigns subjects to one of two arms using a uniform distribution.

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Allocation Ratio|Design Card|$AllocRatio$|

Steps:

1. Generate a random number $u$ from a $Uniform(0,1)$ distribution.
2. Define $p$ as the allocation fraction for the control arm, with $1 - p$ as the allocation fraction for the treatment arm.
3. If $u \leq p$, assign the subject to the control arm; otherwise, assign them to the treatment arm.
4. Ensure that the total sample size equals the sum of subjects in the control and treatment arms.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Randomization" = c(
            "Load allocation ratio", 
            "Convert allocation ratios to fractions",
            "Calculate sample size per arm",
            "Loop through patients",
            "Assign subjects to arms using uniform distribution",
            "Return allocation indices"
        )
    )
)
```

# Example 2 - Randomize Subjects Using Sample Function

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizationSubjectsUsingSampleFunctionInR.R" class="alert-link">RandomizationSubjectsUsingSampleFunctionInR.R</a></p>
</div>

The R function *RandomizationSubjectsUsingSampleFunctionInR* uses the `sample()` function in R to randomly assign patients to the control and treatment arms.

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Allocation Ratio|Design Card|$AllocRatio$|

Steps:

1. Let $p$ be the allocation fraction for the control arm, with $1 - p$ as the allocation fraction for the treatment arm.
2. Compute the expected sample sizes (rounded) for both arms using the allocation fraction and total sample size.
3. Randomly assign subject indices to the control and treatment arms using the `sample()` function in R.
4. Create a vector of zeros with size equal to the total number of subjects ($NumSub$), then replace the zeros with ones for indices corresponding to the treatment arm.

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Randomization" = c(
            "Load allocation ratio", 
            "Convert allocation ratios to fractions",
            "Calculate sample size per arm",
            "Loop through patients",
            "Assign subjects to arms using sample function",
            "Return allocation indices"
        )
    )
)
```

# Example 3 - Randomize Subjects Using randomizeR Package

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/BlockRandomizationSubjectsUsingRPackage.R" class="alert-link">BlockRandomizationSubjectsUsingRPackage.R</a></p>
</div>

The function *BlockRandomizationSubjectsUsingRPackage.R* uses the `randomizeR::pbrPar()` function from the **randomizeR** package to perform block randomization.

In small-sample studies, imbalances between groups can be minimized by using restricted randomization. This approach ensures that treatment groups are allocated in the desired proportions while maintaining randomness.

The **permuted block randomization** technique divides study participants into blocks, within which subjects are randomly assigned to treatment groups. This method ensures that the desired allocation proportions are exactly maintained within each block while keeping the order of assignments random.

Refer to the table below for the definitions of the user-defined parameters used in this example.

|**User parameter**|**Definition**|
|---|------|
|**BlockSize1**|Size of the first block for the randomization process.|
|**BlockSize2**|Size of the second block for the randomization process.|
|*...*|Block sizes for additional blocks follow the same format, incrementing the block number (e.g., BlockSize3, BlockSize4, etc.).|

This function also uses parameters specified in East Horizon. Refer to the tables below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Allocation Ratio|Design Card|$AllocRatio$|

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Randomization" = c(
            "Load allocation ratio and block sizes", 
            "Loop through patients",
            "Assign subjects to arms using block randomization",
            "Return allocation indices"
        )
    )
)
```

# Example 4 - Randomize Subjects Across Multiple Arms

<div class="alert alert-primary" role="alert">
  <p style="margin-bottom:0"> This example is related to this R file: <a href="https://github.com/Cytel-Inc/CyneRgy/blob/main/inst/Examples/RandomizeSubjects/R/RandomizeSubjectsAcrossMultipleArms.R" class="alert-link">RandomizeSubjectsAcrossMultipleArms.R</a></p>
</div>

The function *RandomizeSubjectsAcrossMultipleArms* demonstrates how to perform randomization when a clinical trial involves more than two arms, such as a control group and multiple experimental treatment groups.

This function dynamically handles any number of arms and allocates subjects according to specified allocation ratios. It extends the logic of the two-arm randomization examples to generalize across an arbitrary number of treatment arms.

The function first converts allocation ratios into fractions summing to one, then determines the number of subjects per arm according to these proportions. Any remaining subjects (due to rounding) are distributed to the arms with the largest fractional parts to ensure that the total sample size is preserved. A treatment ID vector is then created (0 for control, 1...N for treatment arms), randomly shuffled, and returned.

This function does not use any user-defined parameters, but uses parameters specified in East Horizon. Refer to the table below for more information.

|**Name of the parameter in East Horizon**|**Where to find the parameter in East Horizon**|**Name of the variable in the R script**|
|-----|-----|---|
|Allocation Ratio for Control Arm|Design Card|$AllocRatio[ 1 ]$|
|Allocation Ratio for Arm 1|Design Card|$AllocRatio[ 2 ]$|
|Allocation Ratio for Arm 2|Design Card|$AllocRatio[ 3 ]$|
|...|

The figure below illustrates where this example fits within the R integration points of Cytel products, accompanied by a flowchart outlining the general steps performed by the R code.

```{r echo=FALSE,  warning=FALSE, fig.retina=3}
CyneRgy::PlotExampleFlowchart(
    lIntPoints = list(
        "Randomization" = c(
            "Load total subjects and number of arms",
            "Convert allocation ratios to fractions",
            "Compute sample size per arm",
            "Allocate remaining subjects",
            "Generate treatment ID vector (0...NumArms-1)",
            "Shuffle assignments randomly",
            "Return allocation indices"
        )
    )
)
```
